---
title: "Assignment 2: Coding Basics"
author: Caroline Rowley
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq(1, 100, 4) #I am asking R to create a sequence from 1 to 100 by 4
one_to_100_by_4 <- seq(1, 100, 4) #I am assigning the sequence a name

#2
mean_one_to_100_by_4 <- mean(one_to_100_by_4) #I am asking R to calculate the mean of the data set, and also assigning the mean the name mean_one_to_100_by_4
median_one_to_100_by_4 <- median(one_to_100_by_4) #I am asking R to calculate the median of the data set, and also assigning the median the name median_one_to_100_by_4

#3. 
mean_one_to_100_by_4>median_one_to_100_by_4 #I am asking R to determine if the mean is greater than the median


```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5
gibbon_names <- c("hoolock", "hylobates", "nomascus", "symphalangus") #the names of the students, each representing their genus of gibbon, this is character
scores <- c(100, 65, 87, 26) #scores that each gibbon got on the test
pass <- c(TRUE, TRUE, TRUE, FALSE) #whether or not they passed the test (received a score greater than 50)

#6
gibbon_names is a character vector
scores is a numeric vector
pass is a logical vector

#7
gibbon_test_scores <- data.frame(gibbon_names, scores, pass) #this code creates a data frame, using the vectors as columns in the order listed

#8
colnames(gibbon_test_scores) <- c("gibbon.name", "gibbon.test.score", "gibbon.pass.fail" )
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The data frame allows heterogenous elements such as characters and numbers. A matrix only allows homogenous data to be stored- either all numerical or all characters. A matrix also has a fixed number of rows and columns, whereas a data frame can contain multiple data types in multiple columns called fields.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
ifelse(scores>50, "pass", "fail") #this code will test with if the scores in the data frame are passing (greater than 50).

if(gibbon_test_scores$gibbon.pass.fail>50)
 (print("pass"))
  else(print("fail"))

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:'ifelse' worked for me. I believe this is because R was given two clear options that worked with the data (pass or fail). When I tried to run the 'if' code I got errors over commas that I did not get when running the 'ifelse' code. When I tried to run the if/else (I left the code in the chunk) it had an error that the condition has a length > 1. I tried running this line of code to the data frame, and also directly to the scores vector without success. It is possible that if/else should work, but is not due to user error and my limited understanding of R.
